Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-04 paper/main.tex
fixed table

%\manuel{}{The above replaces the text below?}
%
%In Section~\ref{sec:preliminaries} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion. 
%
%
%In Section~\ref{sec:permutations} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion. 
% 
%As a summary of Algorithm~\ref{alg:umm}, we model rankings in our UMM model, generate a ranking $\sigma$, then invert this ranking to obtain an ordering in line~\ref{line:inversion} and evaluate the ordering $\sigma^{-1}$ in the objective function. This might seem an implementation detail but for the sake of consistency with the literature, we will expect objective functions to evaluate ordering. A natural question is what happens if we did not invert the ranking and evaluated a ranking with an optimization for orderings. For the anxious reader, we advance that the result is similar to the original version of CEGO. 
%
%In CEGO, the objects modeled and evaluated are the same. They claim that they use the swap distance but they are using the \ken  distance, meaning that the object model is a ranking. \ekhine{}{esto es demasido rude?}
%On the function evaluation the authors interpert their permutation as an ordering, as usual. 
%However, there is no inversion between the objects sampled from the model and the objects evaluated. 
%This means that both are rankings (or both are orderings). A natural question is what happens if we did invert the ranking. We have carried out all the experimentation with this modification of CEGO and we have denote this version CEGO\_inv. The original version of CEGO is denoted CEGO\_orig. 
%

@@

Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
\begin{itemize}
\item UMM\_100, UMM\_200 and UMM\_400 are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM\_100, UMM\_200 wont reach the right corner of the figure. As mentioned in Sec.~\ref{sec:sampling}, the behavior of
UMM depends on the value of $m$. 
\item CEGO\_orig and CEGO\_inv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
\end{itemize}

@@

%
%\paragraph{Figure description} Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
%\begin{itemize}
%\item UMM\_100, UMM\_200 and UMM\_400 are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM\_100, UMM\_200 wont reach the right corner of the figure. 
%\item CEGO\_orig and CEGO\_inv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
%\end{itemize}
%
%\paragraph{Analysis} We start our analysis with the three version of UMM. We can see a general trend for all the instances: After the 10 random initial permutations, the fitness of the new solutions decreases with a linear trend along iterations. The (close to) linear decrease is given by the linear decrease in the expected distance of the model which is sampled at each step, described in Section~\ref{sec:sampling}. Overall, the improvement does not seem to converge not even for the longer runs of 400 evaluations.
%The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations. 
%
%
%The red and the gold lines correspond to CEGO, the original and our modified versions. For the reason for the abrupt change between both version for every instance, we refer the reader to Section~\ref{sec:exper_ro}. For the original version, there is a global decrease in the fitness function. However, a closer analysis of each iteration does not show a real convergence to any point. However, with the modified version in which rankings are modeled and orderings are evaluated we can appreciate the a decrease in terms of fitness per evaluation. Overall, the modified CEGO has a very abrupt decrease at the first 50 iterations. At one point, the algorithm seems to have converged and very little improvement is don from the 150-th iteration onwards.  The average running time is approximately 22 hours for the 400 function evaluations, which is 300 times slower than UMM for the same number of function evaluations. 
%
%The performance of the UMM is clearly better than the performance of the original CEGO for the same problem. For CEGO\_inv on the other hand improves the results of the UMM with the current experimental setting. However, UMM seems to be improving the quality of the solutions until the last iteration while CEGO\_inv seems to be stucked at a local optimum. Further analysis lines will consider a larger budget of fitness evaluation. 
%
%Moreover, UMM seems to scale good for larger instances in quality performance. Note that the instances are order from top to bottom in the size of the permutations. 
%The computational complexity of UMM is $O(n \log n)$ at each iteration while both versions of CEGO need \manuel{tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.
%

@@

\begin{table*}[tb]%time
\begin{tabular}{llrrrrr}
\toprule
Problem &    Instance &  CEGO\_inv &  CEGO\_orig &  UMM\_400 &   UMM\_200 &  UMM\_100 \\
\midrule
    LOP &    N-p40-01 &   1195.90 &    1170.81 &     3.12 &  0.850000 &     0.21 \\
    LOP &    N-p40-02 &   1207.24 &    1183.09 &     3.11 &  0.830000 &     0.20 \\
    LOP &  N-t59b11xx &   1265.98 &    1240.36 &     3.78 &  0.950000 &     0.23 \\
    LOP &  N-t59d11xx &   1269.93 &    1252.89 &     3.93 &  0.966667 &     0.25 \\
    LOP &  N-sgb75.01 &   1994.22 &    2004.70 &    10.88 &  2.730000 &     0.72 \\
    LOP &  N-sgb75.02 &   1995.09 &    1975.41 &    10.88 &  2.760000 &     0.69 \\
   PFSP &       rec05 &   1099.63 &     917.81 &     0.80 &  0.200000 &     0.10 \\
   PFSP &       rec13 &   1109.51 &    1043.08 &     0.85 &  0.200000 &     0.10 \\
   PFSP &       rec19 &   1090.06 &    1087.38 &     1.76 &  0.470000 &     0.10 \\
   PFSP &       rec31 &   1387.09 &    1367.26 &     4.77 &  1.280000 &     0.32 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[tb]%fitness
\begin{tabular}{llrrrrr}
\toprule
Problem &    Instance &   CEGO\_inv &  CEGO\_orig &    UMM\_400 &       UMM\_200 &    UMM\_100 \\
\midrule
    LOP &    N-p40-01 &    10496.6 &    15027.7 &    12453.7 &  1.336170e+04 &    14649.9 \\
    LOP &    N-p40-02 &    11714.7 &    15150.2 &    13191.9 &  1.420900e+04 &    15206.2 \\
    LOP &  N-t59b11xx &   102408.0 &   126204.0 &   114726.5 &  1.202020e+05 &   132039.5 \\
    LOP &  N-t59d11xx &    12692.2 &    30199.8 &    16775.4 &  2.352089e+04 &    29884.1 \\
    LOP &  N-sgb75.01 &  1893533.2 &  2095899.6 &  1983521.2 &  2.004006e+06 &  2046895.1 \\
    LOP &  N-sgb75.02 &  1895801.8 &  2069286.9 &  1973083.9 &  2.001244e+06 &  2039356.1 \\
   PFSP &       rec05 &     1276.2 &     1318.7 &     1282.1 &  1.287000e+03 &     1320.2 \\
   PFSP &       rec13 &     1992.9 &     2096.0 &     2077.7 &  2.108500e+03 &     2131.6 \\
   PFSP &       rec19 &     2264.5 &     2328.4 &     2318.5 &  2.352100e+03 &     2386.5 \\
   PFSP &       rec31 &     3415.5 &     3512.3 &     3420.5 &  3.490900e+03 &     3474.3 \\
\bottomrule
\end{tabular}\end{table*}

>>>>

%\manuel{}{The above replaces the text below?}
%
%In Section~\ref{sec:preliminaries} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion. 
%
%
%In Section~\ref{sec:permutations} we highlighted the great difference of rankings and orderings, usually ommitted in the literature. In this lines, we show the consequences of misinterpreting permutations. We remind the reader that statistical models (Borda or the \ken distance) are defined for rankings while standard literature in optimization considers orderings. This is the reason while in the optimization field we will usually find the Swap distance instead of the \ken distance. However, both are usually claimed to be the same, which is false: they are the same in particular cases, for example when  one of the permutations is the identity or an inversion. 
% 
%As a summary of Algorithm~\ref{alg:umm}, we model rankings in our UMM model, generate a ranking $\sigma$, then invert this ranking to obtain an ordering in line~\ref{line:inversion} and evaluate the ordering $\sigma^{-1}$ in the objective function. This might seem an implementation detail but for the sake of consistency with the literature, we will expect objective functions to evaluate ordering. A natural question is what happens if we did not invert the ranking and evaluated a ranking with an optimization for orderings. For the anxious reader, we advance that the result is similar to the original version of CEGO. 
%
%In CEGO, the objects modeled and evaluated are the same. They claim that they use the swap distance but they are using the \ken  distance, meaning that the object model is a ranking. \ekhine{}{esto es demasido rude?}
%On the function evaluation the authors interpert their permutation as an ordering, as usual. 
%However, there is no inversion between the objects sampled from the model and the objects evaluated. 
%This means that both are rankings (or both are orderings). A natural question is what happens if we did invert the ranking. We have carried out all the experimentation with this modification of CEGO and we have denote this version \CEGOinv. The original version of CEGO is denoted \CEGOorig. 
%

@@

Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
\begin{itemize}
\item UMM$_{100}$, UMM$_{200}$ and UMM$_{400}$ are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM$_{100}$, UMM$_{200}$ wont reach the right corner of the figure. As mentioned in Sec.~\ref{sec:sampling}, the behavior of
UMM depends on the value of $m$. 
\item \CEGOorig and \CEGOinv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
\end{itemize}

@@

%
%\paragraph{Figure description} Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
%\begin{itemize}
%\item UMM$_{100}$, UMM$_{200}$ and UMM$_{400}$ are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM$_{100}$, UMM$_{200}$ wont reach the right corner of the figure. 
%\item \CEGOorig and \CEGOinv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
%\end{itemize}
%
%\paragraph{Analysis} We start our analysis with the three version of UMM. We can see a general trend for all the instances: After the 10 random initial permutations, the fitness of the new solutions decreases with a linear trend along iterations. The (close to) linear decrease is given by the linear decrease in the expected distance of the model which is sampled at each step, described in Section~\ref{sec:sampling}. Overall, the improvement does not seem to converge not even for the longer runs of 400 evaluations.
%The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations. 
%
%
%The red and the gold lines correspond to CEGO, the original and our modified versions. For the reason for the abrupt change between both version for every instance, we refer the reader to Section~\ref{sec:exper_ro}. For the original version, there is a global decrease in the fitness function. However, a closer analysis of each iteration does not show a real convergence to any point. However, with the modified version in which rankings are modeled and orderings are evaluated we can appreciate the a decrease in terms of fitness per evaluation. Overall, the modified CEGO has a very abrupt decrease at the first 50 iterations. At one point, the algorithm seems to have converged and very little improvement is don from the 150-th iteration onwards.  The average running time is approximately 22 hours for the 400 function evaluations, which is 300 times slower than UMM for the same number of function evaluations. 
%
%The performance of the UMM is clearly better than the performance of the original CEGO for the same problem. For \CEGOinv on the other hand improves the results of the UMM with the current experimental setting. However, UMM seems to be improving the quality of the solutions until the last iteration while \CEGOinv seems to be stucked at a local optimum. Further analysis lines will consider a larger budget of fitness evaluation. 
%
%Moreover, UMM seems to scale good for larger instances in quality performance. Note that the instances are order from top to bottom in the size of the permutations. 
%The computational complexity of UMM is $O(n \log n)$ at each iteration while both versions of CEGO need \manuel{tu sabes}{?}. This difference is evident in the time performance of the algorithms. The details can be seen in Table~\ref{table:time}.
%

@@

\begin{table*}[tb]%time
  \caption{Mean fitness (and standard deviation), over 10 independent runs for each problem instance, of the best solution found at the end of each run of the original CEGO (\CEGOorig) and its improved version (\CEGOinv) and three configurations of UMM that different in the budget of evaluations $m \in \{100, 200, 400\}$. The runs of CEGO used a budget of $m=400$.\label{tab:results_fitness}}
\begin{tabular}{llrrrrr}
\toprule
\bf Problem & \bf   Instance &  \CEGOinv &  \CEGOorig &  UMM$_{400}$ &   UMM$_{200}$ &  UMM$_{100}$ \\
\midrule
    LOP &    N-p40-01 &   1195.90 &    1170.81 &     3.12 &  0.850000 &     0.21 \\
    LOP &    N-p40-02 &   1207.24 &    1183.09 &     3.11 &  0.830000 &     0.20 \\
    LOP &  N-t59b11xx &   1265.98 &    1240.36 &     3.78 &  0.950000 &     0.23 \\
    LOP &  N-t59d11xx &   1269.93 &    1252.89 &     3.93 &  0.966667 &     0.25 \\
    LOP &  N-sgb75.01 &   1994.22 &    2004.70 &    10.88 &  2.730000 &     0.72 \\
    LOP &  N-sgb75.02 &   1995.09 &    1975.41 &    10.88 &  2.760000 &     0.69 \\
   PFSP &       rec05 &   1099.63 &     917.81 &     0.80 &  0.200000 &     0.10 \\
   PFSP &       rec13 &   1109.51 &    1043.08 &     0.85 &  0.200000 &     0.10 \\
   PFSP &       rec19 &   1090.06 &    1087.38 &     1.76 &  0.470000 &     0.10 \\
   PFSP &       rec31 &   1387.09 &    1367.26 &     4.77 &  1.280000 &     0.32 \\
\bottomrule
\end{tabular}
\end{table*}

\begin{table*}[tb]%fitness
\begin{tabular}{llrrrrr}
\toprule
Problem &    Instance &   \CEGOinv &  \CEGOorig &    UMM$_{400}$ &       UMM$_{200}$ &    UMM$_{100}$ \\
\midrule
    LOP &    N-p40-01 &    10496.6 &    15027.7 &    12453.7 &  1.336170e+04 &    14649.9 \\
    LOP &    N-p40-02 &    11714.7 &    15150.2 &    13191.9 &  1.420900e+04 &    15206.2 \\
    LOP &  N-t59b11xx &   102408.0 &   126204.0 &   114726.5 &  1.202020e+05 &   132039.5 \\
    LOP &  N-t59d11xx &    12692.2 &    30199.8 &    16775.4 &  2.352089e+04 &    29884.1 \\
    LOP &  N-sgb75.01 &  1893533.2 &  2095899.6 &  1983521.2 &  2.004006e+06 &  2046895.1 \\
    LOP &  N-sgb75.02 &  1895801.8 &  2069286.9 &  1973083.9 &  2.001244e+06 &  2039356.1 \\
   PFSP &       rec05 &     1276.2 &     1318.7 &     1282.1 &  1.287000e+03 &     1320.2 \\
   PFSP &       rec13 &     1992.9 &     2096.0 &     2077.7 &  2.108500e+03 &     2131.6 \\
   PFSP &       rec19 &     2264.5 &     2328.4 &     2318.5 &  2.352100e+03 &     2386.5 \\
   PFSP &       rec31 &     3415.5 &     3512.3 &     3420.5 &  3.490900e+03 &     3474.3 \\
\bottomrule
\end{tabular}\end{table*}
